SylvanSeeds, a seed germination database for temperate deciduous forests

### 2020-10-12 README.Rmd
readme pic
small readme change
accepted ms 🌳🍋🌱

![SylvanSeeds](data/2017070313 Blue Ridge Parkway small.JPG)

@@

This repository stores all the information related to the publication *SylvanSeeds, a seed germination database for temperate deciduous forests*, including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the manuscript. It also stores the code of the *SylvanSeeds* `shiny` app, that can be used to visualize the database.

@@

* `data` Original raw files including the primary result of the search in the Web of Science, and search notes.
* `doc` Files to create the manuscript using `Rmarkdown`.
* `results` Output of the `R` scripts, including the clean database and files needed to create the manuscript.
* `src` Scripts in `R` language used to clean the raw data, perform the analyses of the manuscript, and build the `SylvanSeeds` app.

@@

> Fernández-Pascual, E (2020) SylvanSeeds, a seed germination database for temperate deciduous forests. Journal of Vegetation Science, in press.

>>>>

![SylvanSeeds](data/2017070313 Blue Ridge Parkway small.jpg)

@@

This repository stores all the information related to the publication [*SylvanSeeds, a seed germination database for temperate deciduous forests*](https://doi.org/10.1111/jvs.12960), including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the [manuscript](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/manuscript.md). It also stores the code of the *SylvanSeeds* `shiny` app, that can be used to visualize the database.

@@

* `data` Original raw files including the primary result of the search in the Web of Science, and search notes.
* `doc` Files to create the [manuscript](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/manuscript.md) using `Rmarkdown`.
* `results` Output of the `R` scripts, including the clean database and files needed to create the manuscript.
* `src` Scripts in `R` language used to clean the raw data, perform the analyses of the manuscript, and build the `SylvanSeeds` app.

@@

> Fernández-Pascual, E. (2020). [SylvanSeeds, a seed germination database for temperate deciduous forests.](https://doi.org/10.1111/jvs.12960) *Journal of Vegetation Science*

### 2020-10-12 doc/manuscript.Rmd
accepted ms 🌳🍋🌱

---
title: "SylvanSeeds, a seed germination database for temperate deciduous forests"
subtitle: "Running title: SylvanSeeds, a forest germination database"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

A systematic approach was followed to search for data, with the purpose of prioritising work and maximising its utility for vegetation science. In short, this approach consisted in (i) obtaining a representative number of vegetation plots for the biome from sPlot (https://www.idiv.de/en/splot.html); and (ii) using the plots to produce a list of frequent species (i.e. species that occur in at least 5 % of the plots of a region).

@@

The list of frequent species (plus the synonyms recorded in the relevés) was incorporated into a Boolean search string, together with the words "(seed OR seeds) AND (dormancy OR germination)”. This string was searched in the *Thompson Reuters Web of Science* on 5 Mar 2019, returning 6,791 results. The processing of these results is described in a PRISMA flow diagram (**Appendix S1**). A first filter of the results by the relevance of the title retained 1,489 references. An effort was made to access the full text of all these references, using the following channels: (i) “Find Full Text” functionality in *EndNote X9*, (ii) online access to the journals using the institutional subscriptions of the University of Oviedo; (iii) request to the authors through *ResearchGate*; (iv) general search of *pdf* files in *Google* using the article details. Unfortunately, 236 references could not be accessed. Most of these belonged to smaller publishers not included in the subscriptions of the University of Oviedo, and to older journal issues which had not been digitalised. The resources of this project did not make it possible to further pursue these references, but they are included as a *BibTeX* file (**Appendix S2**) in case they can be processed in the future. The found texts were accessed to retrieve data for the database. At this stage, references were considered to be non-relevant if they (i) did not pertain to species in the original frequent species list; (ii) did not study germination in species monocultures (e.g. competition studies); (iii) did not report the minimal information needed (number of seeds sown, germination percentage and germination temperature). Records were also excluded when they pertained to special germination treatments that do not allow for general comparisons (i.e. use of germination promoters such as the plant hormone gibberellic acid; drought treatments, including polyethylene glycol solutions; application of heavy metals or other pollutants; smoke-related germination promoters such as karrikins). Of the accessed references, 698 were non-relevant (*BiBTeX* file in **Appendix S3**) and 555 provided relevant data (*BiBTeX* file in **Appendix S4**).

The systematic approach applied may fail to capture references on species belonging to the nemoral biome for three reasons: (1) the species is not recorded in the vegetation database; (2) the species has been recorded, but it is infrequent (see above the frequency criterion); or (3) the species is recorded and frequent and goes into the list, but the relevant germination references are not found by the web search because the species name is not mentioned in the title or the abstract. While points 1 and 2 are intentionally excluded by the applied methodology, point 3 would be a failure of the methodology (specifically, a failure of the search string to find a relevant paper). For this reason, a protocol was applied to detect the importance of point 3, using the book of Baskin and Baskin [-@RN3214] as a reference. After completing the database, species names in the frequent species list for which no relevant papers had been found were searched in the taxonomical index of the Baskin and Baskin book. There were 273 species with missing data and mentioned by Baskin and Baskin in their index. A random sample of 10 % of these species were selected, and their references as mentioned in the book were searched. All of them were found to be either inaccessible (national journals or old conference proceedings not available online) or to have no relevant data (i.e. provided no primary germination data that could be retrieved). From this trial, it was concluded that the search string had found a reasonable amount of the available information, by comparing it to the best reference available, the review book of Baskin and Baskin.

@@

The final database contains `r MSrecords` records (germination proportions for a given seed lot of a species, recorded in a set of experimental conditions) from 555 references. The plant materials had been collected across the Temperate Broadleaf and Mixed Forest biome and surrounding areas (**Fig. \@ref(fig:fig1)**), in both hemispheres. Although the frequent species list was created using only relevés from the northern hemisphere, the web search of germination articles found studies of those species that had been performed in the southern hemisphere. These records are included in the database, and their geographical origin is reported. The oldest record was from 1920 and the top three contributing countries were the USA (`r MScountries[1, 2]`), the UK (`r MScountries[2, 2]`) and Japan (`r MScountries[3, 2]`). There were `r MSspecies` species represented, from `r MSfamilies` seed plant families. The total estimate of seeds used in the experiments was `r MSseeds`. The range of germination temperatures (weighted average of the daily thermoperiod) went from `r MSminT` to `r MSmaxT` ºC, with `r MSaltY` records of constant temperatures and `r MSaltN` of alternating temperatures. Light was used in `r MSlightY` records, darkness in `r MSlightN` and `r MSlightNA` did not provide information on this parameter. The experiments were performed with unstratified seeds in `r MSstratN` records, and of the rest, the majority (`r MSstratCold`) went through cold stratification. Scarification was applied to `r MSscarY` records.

@@

The database is provided as a *csv* file, comma separated, named **Appendix S5** (see online Supporting Information). The first row of the file contains the header data, with the following variables: *Taxon* (the original species names, as used in the references), *TPLName* (*The Plant List* species names), *Family* (the botanical family of the species), *Reference* (bibliographic source of the record), *Year* (year the seed lot was collected), *Country* (country where the seed lot was collected), *Population* (geographical information of the seed lot), *Latitude* (approximated latitude where the seed lot was collected, in decimal degrees), *Longitude* (approximated longitude where the seed lot was collected, in decimal degrees), *Dry_storage* (binary variable indicating whether the seed lot was kept at dry storage for more than a month between seed collection and start of the experiments), *Setting* (environs used to control germination temperature: either a greenhouse, a climatically-controlled incubator, or room temperature), *Sterilization* (note on surface sterilization treatments applied to the seed lot before the experiments), *Nutrients* (note on nutrient application to the germination substrate), *Substrate* (substrate used to hold the seeds during the germination incubation), *Container_type* (type of container used to hold the substrate and the seeds during the experiments), *Container_size* (dimensions of the germination containers), *Container_number* (number of germination containers used for the treatment), *Number_per_container* (number of seeds sown in each container), *Incubation_days* (number of days between the start of the experiment, not including stratification, and the last day when germinated seeds were counted), *Scarification* (binary variable indicating whether the seed lot was scarified before the test), *Stratification_days* (number of days the seed lot was exposed to any type of stratification, before the test), *Stratification_temperature* (temperature or temperatures used during the stratification treatment, in degrees Celsius), *Stratification_type* (type of stratification, which can be none, cold, warm or combinations of cold and warm), *Stratification* (binary variable indicating whether the seed lot was stratified or not before the test), *Light* (binary variable indicating whether the seed lot was germinated in light or in darkness), *Photoperiod* (number of hours of exposure to light in the daily photoperiod), *Alternating* (binary variable indicating whether the germination test was conducted under constant or alternating temperatures), *Tdif* (difference in degrees between the hottest and the coldest temperatures of the daily thermoperiod), *Tmax* (hottest temperature in the daily thermoperiod), *Tmin* (coldest temperature in the daily thermoperiod), *Tmean* (mean germination temperature, weighted by the length of each phase of the daily thermoperiod), *Temperature* (aggregation of the mean germination temperature in 5 ºC intervals), *Germinated* (count of seeds that germinated during the experiment), *Number_seeds* (count of seeds used in the experiment). Each row below the header represents a record for a seed lot germinated in a given set of experimental conditions. Presenting the response variable as a proportion (i.e. number of successes *Germinated*, number of trials *Number_seeds*) has important benefits for meta-analysis of the database, as proportions can be analysed directly using a GLM, and a variance term does not need to be specified since it is incorporated as part of the analysis [@RN4751].

@@

To facilitate the visualization of the database, the *SylvanSeeds* app was written using the *shiny* package [@RN4661]. It is publicly accessible at https://efernandezpascual.shinyapps.io/sylvanseeds/. The app uses the *tidyverse* package [@RN4662] to aggregate and show results for species and experimental treatments (i.e. aggregating all seed lots of the same species germinated in the same experimental conditions). To facilitate comparisons, germination temperatures are aggregated to 5 ºC intervals. When there is only one seed lot per species and combination of experimental conditions, the binomial 95% confidence interval is calculated using the Wilson method in the *binom* package [@RN2964]. When there is more than one seed lot per species and combination of experimental conditions, the aggregate proportion and binomial confidence intervals are calculated using binomial-normal meta-analysis models [@RN4663] as implemented in the package *metafor* [@RN2274]. By visiting the app, users can consult the available germination information for a species (**Fig. \@ref(fig:fig2)**), the origin of its seed lots, and the bibliographical references for the species.

@@

@@

**Appendix S1** PRISMA 2009 flow chart describing the process of reference screening and inclusion. **Appendix S2** Bibliography in BibTeX format listing the not-found full texts.  **Appendix S3** Bibliography in BibTeX format listing the non-relevant references.  **Appendix S4** Bibliography in BibTeX format listing the relevant references included in the *SylvanSeeds* database.  **Appendix S5** *SylvanSeeds* database in a single *.csv* file.

All persons can use the database providing they cite this paper properly in any publications or in the metadata of any derived products that are produced using the database. The database is provided as **Appendix S5** of this manuscript, but please do visit https://efernandezpascual.github.io/home/sylvanseeds.html to check for updates. It can be visualized with the *SylvanSeeds* app at  https://efernandezpascual.shinyapps.io/sylvanseeds/. The code to create this manuscript and the app is stored at https://github.com/efernandezpascual/sylvanseeds.

# Figures

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Geographical distribution of the germination records in the database. Each golden circle is a record. The green areas correspond to the extension of the Temperate Broadleaf and Mixed Forests biome according to the *WWF*.", fig.width=7, fig.height=3}
p1
```

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Example of the germination records as shown by the *SylvanSeeds* app. Records for one species, the European common beech, *Fagus sylvatica*. Each panel shows the results for a combination of experimental conditions, with the germination temperature varying within each panel. Bars represent the mean germination proportion and brackets the 95% binomial confidence interval.", fig.width=7, fig.height=5}
p2
```

>>>>

---
title: "SylvanSeeds, a seed germination database for temperate deciduous forests"
output: github_document
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

A systematic approach was followed to search for data, with the purpose of prioritising work and maximising its utility for vegetation science. In short, this approach consisted in (i) obtaining a representative number of vegetation plots for the biome from sPlot (https://www.idiv.de/en/splot.html); and (ii) using the plots to produce a list of frequent species (i.e. species that occur in at least 5% of the plots of a region).

@@

The list of frequent species (plus the synonyms recorded in the relevés) was incorporated into a Boolean search string, together with the words "(seed OR seeds) AND (dormancy OR germination)”. This string was searched in the *Thompson Reuters Web of Science* on 5 Mar 2019, returning 6,791 results. The processing of these results is described in a PRISMA flow diagram [(**Appendix S1**)](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendices%20S1-S4.pdf). A first filter of the results by the relevance of the title retained 1,489 references. An effort was made to access the full text of all these references, using the following channels: (i) “Find Full Text” functionality in *EndNote X9*, (ii) online access to the journals using the institutional subscriptions of the University of Oviedo; (iii) request to the authors through *ResearchGate*; (iv) general search of *pdf* files in *Google* using the article details. Unfortunately, 236 references could not be accessed. Most of these belonged to smaller publishers not included in the subscriptions of the University of Oviedo, and to older journal issues which had not been digitalised. The resources of this project did not make it possible to further pursue these references, but they are included as a *BibTeX* file [(**Appendix S2**)](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendices%20S1-S4.pdf) in case they can be processed in the future. The found texts were accessed to retrieve data for the database. At this stage, references were considered to be non-relevant if they (i) did not pertain to species in the original frequent species list; (ii) did not study germination in species monocultures (e.g. competition studies); (iii) did not report the minimal information needed (number of seeds sown, germination percentage and germination temperature). Records were also excluded when they pertained to special germination treatments that do not allow for general comparisons (i.e. use of germination promoters such as the plant hormone gibberellic acid; drought treatments, including polyethylene glycol solutions; application of heavy metals or other pollutants; smoke-related germination promoters such as karrikins). Of the accessed references, 698 were non-relevant (*BiBTeX* file in [**Appendix S3**](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendices%20S1-S4.pdf)) and 555 provided relevant data (*BiBTeX* file in [**Appendix S4**](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendices%20S1-S4.pdf)).

The systematic approach applied may fail to capture references on species belonging to the nemoral biome for three reasons: (1) the species is not recorded in the vegetation database; (2) the species has been recorded, but it is infrequent (see above the frequency criterion); or (3) the species is recorded and frequent and goes into the list, but the relevant germination references are not found by the web search because the species name is not mentioned in the title or the abstract. While points 1 and 2 are intentionally excluded by the applied methodology, point 3 would be a failure of the methodology (specifically, a failure of the search string to find a relevant paper). For this reason, a protocol was applied to detect the importance of point 3, using the book of Baskin and Baskin [-@RN3214] as a reference. After completing the database, species names in the frequent species list for which no relevant papers had been found were searched in the taxonomical index of the Baskin and Baskin book. There were 273 species with missing data and mentioned by Baskin and Baskin in their index. A random sample of 10% of these species were selected, and their references as mentioned in the book were searched. All of them were found to be either inaccessible (national journals or old conference proceedings not available online) or to have no relevant data (i.e. provided no primary germination data that could be retrieved). From this trial, it was concluded that the search string had found a reasonable amount of the available information, by comparing it to the best reference available, the review book of Baskin and Baskin.

@@

The final database contains `r MSrecords` records (germination proportions for a given seed lot of a species, recorded in a set of experimental conditions) from 555 references. The plant materials had been collected across the Temperate Broadleaf and Mixed Forest biome and surrounding areas (**Fig. 1**), in both hemispheres. Although the frequent species list was created using only relevés from the northern hemisphere, the web search of germination articles found studies of those species that had been performed in the southern hemisphere. These records are included in the database, and their geographical origin is reported. The oldest record was from 1920 and the top three contributing countries were the USA (`r MScountries[1, 2]`), the UK (`r MScountries[2, 2]`) and Japan (`r MScountries[3, 2]`). There were `r MSspecies` species represented, from `r MSfamilies` seed plant families. The total estimate of seeds used in the experiments was `r MSseeds`. The range of germination temperatures (weighted average of the daily thermoperiod) went from `r MSminT` to `r MSmaxT` ºC, with `r MSaltY` records of constant temperatures and `r MSaltN` of alternating temperatures. Light was used in `r MSlightY` records, darkness in `r MSlightN` and `r MSlightNA` did not provide information on this parameter. The experiments were performed with unstratified seeds in `r MSstratN` records, and of the rest, the majority (`r MSstratCold`) went through cold stratification. Scarification was applied to `r MSscarY` records.

```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 1** *Geographical distribution of the germination records in the database. Each golden circle is a record. The green areas correspond to the extension of the Temperate Broadleaf and Mixed Forests biome according to the *WWF*.*", fig.width=7, fig.height=3}
knitr::include_graphics("../results/Fig1.png")
```

@@

The database is provided as a *csv* file, comma separated [(**Appendix S5**)](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendix%20S5%20-%20Database.csv). The first row of the file contains the header data, with the following variables: *Taxon* (the original species names, as used in the references), *TPLName* (*The Plant List* species names), *Family* (the botanical family of the species), *Reference* (bibliographic source of the record), *Year* (year the seed lot was collected), *Country* (country where the seed lot was collected), *Population* (geographical information of the seed lot), *Latitude* (approximated latitude where the seed lot was collected, in decimal degrees), *Longitude* (approximated longitude where the seed lot was collected, in decimal degrees), *Dry_storage* (binary variable indicating whether the seed lot was kept at dry storage for more than a month between seed collection and start of the experiments), *Setting* (environs used to control germination temperature: either a greenhouse, a climatically-controlled incubator, or room temperature), *Sterilization* (note on surface sterilization treatments applied to the seed lot before the experiments), *Nutrients* (note on nutrient application to the germination substrate), *Substrate* (substrate used to hold the seeds during the germination incubation), *Container_type* (type of container used to hold the substrate and the seeds during the experiments), *Container_size* (dimensions of the germination containers), *Container_number* (number of germination containers used for the treatment), *Number_per_container* (number of seeds sown in each container), *Incubation_days* (number of days between the start of the experiment, not including stratification, and the last day when germinated seeds were counted), *Scarification* (binary variable indicating whether the seed lot was scarified before the test), *Stratification_days* (number of days the seed lot was exposed to any type of stratification, before the test), *Stratification_temperature* (temperature or temperatures used during the stratification treatment, in degrees Celsius), *Stratification_type* (type of stratification, which can be none, cold, warm or combinations of cold and warm), *Stratification* (binary variable indicating whether the seed lot was stratified or not before the test), *Light* (binary variable indicating whether the seed lot was germinated in light or in darkness), *Photoperiod* (number of hours of exposure to light in the daily photoperiod), *Alternating* (binary variable indicating whether the germination test was conducted under constant or alternating temperatures), *Tdif* (difference in degrees between the hottest and the coldest temperatures of the daily thermoperiod), *Tmax* (hottest temperature in the daily thermoperiod), *Tmin* (coldest temperature in the daily thermoperiod), *Tmean* (mean germination temperature, weighted by the length of each phase of the daily thermoperiod), *Temperature* (aggregation of the mean germination temperature in 5 ºC intervals), *Germinated* (count of seeds that germinated during the experiment), *Number_seeds* (count of seeds used in the experiment). Each row below the header represents a record for a seed lot germinated in a given set of experimental conditions. Presenting the response variable as a proportion (i.e. number of successes *Germinated*, number of trials *Number_seeds*) has important benefits for meta-analysis of the database, as proportions can be analysed directly using a GLM, and a variance term does not need to be specified since it is incorporated as part of the analysis [@RN4751].

@@

To facilitate the visualization of the database, the *SylvanSeeds* app was written using the *shiny* package [@RN4661]. It is publicly accessible at https://efernandezpascual.shinyapps.io/sylvanseeds/. The app uses the *tidyverse* package [@RN4662] to aggregate and show results for species and experimental treatments (i.e. aggregating all seed lots of the same species germinated in the same experimental conditions). To facilitate comparisons, germination temperatures are aggregated to 5 ºC intervals. When there is only one seed lot per species and combination of experimental conditions, the binomial 95% confidence interval is calculated using the Wilson method in the *binom* package [@RN2964]. When there is more than one seed lot per species and combination of experimental conditions, the aggregate proportion and binomial confidence intervals are calculated using binomial-normal meta-analysis models [@RN4663] as implemented in the package *metafor* [@RN2274]. By visiting the app, users can consult the available germination information for a species (**Fig. 2**), the origin of its seed lots, and the bibliographical references for the species.

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "**Figure 2** *Example of the germination records as shown by the SylvanSeeds app. Records for one species, the European common beech, Fagus sylvatica. Each panel shows the results for a combination of experimental conditions, with the germination temperature varying within each panel. Bars represent the mean germination proportion and brackets the 95% binomial confidence interval.*", fig.width=7, fig.height=5}
knitr::include_graphics("../results/Fig2.png")
```

@@

[**Appendices S1-S4**](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendices%20S1-S4.pdf) PRISMA 2009 flow chart describing the process of reference screening and inclusion; bibliography in BibTeX format listing the not-found full texts, the non-relevant references and the relevant references included in the *SylvanSeeds* database.  [**Appendix S5**](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendix%20S5%20-%20Database.csv) *SylvanSeeds* database in a single *.csv* file.

All persons can use the database providing they cite this paper properly in any publications or in the metadata of any derived products that are produced using the database. The database is provided as [**Appendix S5**](https://github.com/efernandezpascual/sylvanseeds/blob/master/doc/Appendix%20S5%20-%20Database.csv) of this manuscript, but please do visit https://efernandezpascual.github.io/home/sylvanseeds.html to check for updates. It can be visualized with the *SylvanSeeds* app at  https://efernandezpascual.shinyapps.io/sylvanseeds/. The code to create this manuscript and the app is stored at https://github.com/efernandezpascual/sylvanseeds.

@@
